---
title: |
  | From DSL to Native in One Process:
  | JIT-Compiling Exo Kernels via xDSL and llvmlite
author: ""
date: ""
output:
  pdf_document:
    latex_engine: pdflatex
    keep_tex: false
header-includes:
  - \usepackage{flushend}
  - \usepackage{titlesec}
  - \usepackage{enumitem}
  - \usepackage{fancyhdr}
  - \usepackage{abstract}
  - \usepackage{booktabs}
  - \usepackage{float}
  - \titleformat{\section}{\normalfont\large\bfseries}{\thesection.}{0.5em}{}
  - \titleformat{\subsection}{\normalfont\normalsize\bfseries}{\thesubsection}{0.5em}{}
  - \titlespacing*{\section}{0pt}{1.8ex plus 0.3ex minus 0.2ex}{0.8ex plus 0.1ex}
  - \titlespacing*{\subsection}{0pt}{1.2ex plus 0.2ex minus 0.1ex}{0.5ex plus 0.1ex}
  - \setlist{nosep, leftmargin=1.2em}
  - \renewcommand{\abstractnamefont}{\normalfont\bfseries}
  - \renewcommand{\abstracttextfont}{\normalfont\small}
  - \setlength{\parskip}{0.5ex plus 0.1ex minus 0.1ex}
  - \setlength{\parindent}{1em}
  - \setlength{\columnsep}{0.3in}
  - \setlength{\abovecaptionskip}{4pt}
  - \setlength{\belowcaptionskip}{-2pt}
  - \pagestyle{fancy}
  - \fancyhf{}
  - \renewcommand{\headrulewidth}{0pt}
  - \fancyfoot[C]{\small\thepage}
classoption: twocolumn
fontsize: 10pt
geometry: "a4paper,left=0.7in,right=0.7in,top=0.75in,bottom=0.75in"
urlcolor: blue
bibliography: []
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE,
                      fig.width=3.4, fig.height=2.4, fig.align="center",
                      out.width="\\columnwidth", fig.pos="H")
library(ggplot2); library(tidyr); library(dplyr); library(forcats)

theme_paper <- theme_minimal(base_size=9) +
  theme(panel.grid.minor=element_blank(),
        plot.title=element_text(size=9, face="bold"),
        axis.title=element_text(size=8), axis.text=element_text(size=7),
        plot.margin=margin(2,2,2,2))

df <- read.csv("../benchmarks/results.csv", stringsAsFactors=FALSE)

# 1D kernels: those whose n column is purely numeric
df_1d <- df %>% filter(grepl("^[0-9]+$", n)) %>% mutate(n=as.numeric(n))
# 2D kernels: matmul, matvec, weighted_sum
df_2d <- df %>% filter(!grepl("^[0-9]+$", n) | kernel %in% c("matmul","matvec"))

# Pivot speedups to long form for faceted plots
sp <- df_1d %>%
  select(kernel, n, exo_speedup, neon_speedup) %>%
  pivot_longer(cols=c(exo_speedup, neon_speedup),
               names_to="variant", values_to="speedup") %>%
  mutate(variant=ifelse(variant=="exo_speedup","Auto-vectorized","NEON intrinsics"),
         speedup_clamped=pmax(speedup, 1.0))

# Summary: peak speedup per kernel per variant
peak <- sp %>% group_by(kernel, variant) %>%
  summarise(peak_speedup=max(speedup), .groups="drop")
```

\begin{abstract}
\noindent
Exo is a user-schedulable language for hardware kernels that only emits C code, requiring a separate compile-link-load cycle. We present a JIT compilation path that lowers Exo's intermediate representation through xDSL (a Python-native MLIR framework) to LLVM IR, then compiles to native code via llvmlite's MCJIT engine, all within a single Python process. We evaluate on three fused multi-pass kernels (softmax, cross-entropy loss, and RMSNorm) which are central to transformer inference and benefit most from kernel fusion. On Apple Silicon, our JIT-compiled fused kernels sustain $4{-}6\times$ speedup over NumPy even at $N{=}4$M, because fusing multiple passes into a single loop eliminates intermediate memory traffic that NumPy cannot avoid. We additionally benchmark eight simpler kernels as baselines. The compiler is approximately 1,250 lines of Python.
\end{abstract}

# Introduction

Writing fast numerical code in Python typically means writing C. The programmer implements a kernel, compiles it to a shared library, loads the result via \texttt{ctypes}, and hopes the pointer types are correct. Changing anything, whether reordering a loop or adjusting a tile size, requires repeating the entire cycle.

Exo [ikarashi2022exo] is a Python-embedded DSL for hardware kernels. The programmer writes a loop nest and transforms it through semantics-preserving scheduling primitives. However, Exo only generates C. The programmer still needs a C compiler, a build system, and an FFI bridge.

We present a JIT compilation path that skips C entirely. Instead of going from Exo to C to clang to a shared object to ctypes, we go from Exo directly to native code in one Python process. We use xDSL [fehr2025xdsl] (Python-native MLIR) to lower Exo's IR to the LLVM dialect, then llvmlite [numba2015llvmlite] (from Numba) to JIT-compile via MCJIT. The result is a callable function pointer, usable with NumPy arrays through a C extension with \texttt{vectorcall} protocol.

We focus our evaluation on three fused multi-pass kernels, namely softmax, cross-entropy loss, and RMSNorm, where kernel fusion provides a sustained advantage over NumPy. We additionally benchmark eight simpler kernels (SAXPY, add, ReLU, dot, embedding, matmul, matvec, weighted sum) as baselines. Each kernel is implemented in two variants: an auto-vectorized version (relying on LLVM's vectorizer) and a hand-written NEON intrinsics version using Exo's \texttt{@instr} mechanism.

# Background

\textbf{Exo} [ikarashi2022exo] is a language for exocompilation. Instead of hiding hardware-specific optimizations inside the compiler, it exposes them as user-level scheduling operations. The programmer writes a loop nest (the algorithm) and transforms it through rewrites (the schedule). Transforms include \texttt{fission}, \texttt{reorder\_loops}, \texttt{divide\_loop}, and \texttt{unroll\_loop}. Each transform preserves semantics. Exo's \texttt{@instr} decorator allows declaring custom hardware intrinsics with scheduling-visible semantics. Exo currently emits C only.

\textbf{xDSL} [fehr2025xdsl] reimplements MLIR's core infrastructure in pure Python, including SSA-based IRs, dialects, regions, and pattern rewriting. It provides the same \texttt{memref}, \texttt{llvm}, and \texttt{builtin} dialect operations as upstream MLIR without requiring a C++ LLVM build.

\textbf{llvmlite} [numba2015llvmlite] provides a Python API for constructing LLVM IR and JIT-compiling via MCJIT. The programmer constructs an \texttt{ir.Module}, parses it into MCJIT, and retrieves a function pointer by name.

# System Design

\textbf{Pipeline Overview.}
The compilation pipeline consists of four stages, each a Python function call, with no subprocesses and no disk I/O:

\noindent
\footnotesize
\begin{verbatim}
 Exo @proc  ->  LoopIR          [Exo]
    -> MLIR (memref+llvm)       [xDSL]
    -> MLIR (llvm only)         [xDSL passes]
    -> LLVM IR -> MCJIT -> fn*  [llvmlite]
    -> Python callable          [C ext]
\end{verbatim}
\normalsize

\textbf{IR Generation.}
The IR generator walks Exo's LoopIR and emits xDSL operations. Expressions become \texttt{llvm.constant}, \texttt{llvm.fadd}, and \texttt{memref.load}. For loops become three basic blocks with an induction variable as a block argument. Allocations become \texttt{llvm.call @malloc} (heap) or \texttt{llvm.alloca} (stack). All floating-point operations carry fast-math flags to enable vectorization.

\textbf{Memory Lowering.}
Standard MLIR lowers \texttt{memref} through descriptors, that is, structs with a base pointer, offset, sizes, and strides. Exo tensors have static shapes and row-major layout, so we bypass descriptors entirely, lowering to direct pointer arithmetic via \texttt{llvm.getelementptr}. Function arguments are passed as bare pointers with \texttt{noalias} attributes, enabling LLVM's auto-vectorizer.

\textbf{Intrinsic Lowering.}
Exo's \texttt{@instr} decorator binds scheduling-visible semantics to hardware operations. We declare NEON intrinsics over a custom \texttt{NEON} memory class representing 128-bit registers (4$\times$f32). The \texttt{ConvertVecIntrinsic} pass lowers these to LLVM vector operations. Available intrinsics include load/store, broadcast, FMA, add, sub, mul, and \texttt{fmax} (via \texttt{llvm.maxnum.v4f32}).

\textbf{JIT Compilation and Caching.}
The final stage translates the xDSL LLVM-dialect module to llvmlite IR and invokes MCJIT with LLVM O2 optimization. The compiled IR is cached on disk, keyed by a hash of all compiler source files. A C extension (\texttt{jitcall.c}) wraps function pointers with Python's \texttt{vectorcall} protocol, extracting buffer pointers from NumPy arrays with approximately 1\,$\mu$s FFI overhead.

# Evaluation

We evaluate on Apple Silicon (AArch64) with ARM NEON 128-bit SIMD. Each benchmark reports throughput in GFLOP/s and speedup over NumPy (which uses Apple's Accelerate framework with AMX for BLAS operations). Timing uses the minimum of 200 \texttt{timeit} repeats.

\textbf{Kernels.}
We organize the 11 benchmarked kernels into three categories. The \emph{fused multi-pass} kernels are our primary focus, as they are where kernel fusion provides a sustained advantage that NumPy cannot match.

The first category is \emph{fused multi-pass} kernels: softmax (max + exp + sum + normalize), cross-entropy (max + exp + sum + log), and RMSNorm (sum-of-squares + scale). NumPy must execute each pass as a separate array operation, materializing intermediates; our JIT kernels fuse all passes into a single loop. The second category is \emph{element-wise} kernels (SAXPY, add, ReLU, dot, embedding), which serve as baselines. Both implementations are memory-bandwidth-bound at large $N$, so speedups converge toward $1\times$. The third category is \emph{matrix} kernels (matmul, matvec, weighted sum), where NumPy's Accelerate/AMX dominates at large sizes.

Each kernel has two JIT variants: \emph{auto-vectorized} (scalar Exo code, LLVM vectorizes) and \emph{NEON intrinsics} (explicit \texttt{neon\_*} instructions via Exo's \texttt{@instr}).

\textbf{Fused Kernels: Sustained Speedup via Fusion.}

```{r fused-plot, fig.width=3.3, fig.height=5.5, out.width="\\columnwidth"}
fused <- sp %>%
  filter(kernel %in% c("softmax","cross_entropy","rmsnorm")) %>%
  mutate(n_fct=factor(n, levels=sort(unique(n))),
         n_lab=ifelse(n>=1e6, paste0(n/1e6,"M"),
                ifelse(n>=1e3, paste0(n/1e3,"K"), as.character(n))),
         kernel=factor(kernel, levels=c("softmax","cross_entropy","rmsnorm"),
                       labels=c("Softmax","Cross-Entropy","RMSNorm")))
n_labs <- fused %>% distinct(n, n_lab) %>% arrange(n)

ggplot(fused, aes(n_fct, speedup_clamped, color=variant, linetype=variant, group=variant)) +
  geom_hline(yintercept=1, color="#bbbbbb", linewidth=0.4) +
  geom_line(linewidth=0.5) +
  geom_point(aes(shape=variant), size=1.0) +
  facet_wrap(~kernel, ncol=1) +
  scale_x_discrete(labels=setNames(n_labs$n_lab, n_labs$n)) +
  scale_color_manual(values=c("Auto-vectorized"="#4C72B0","NEON intrinsics"="#DD8452")) +
  scale_linetype_manual(values=c("Auto-vectorized"="solid","NEON intrinsics"="dashed")) +
  scale_shape_manual(values=c("Auto-vectorized"=16,"NEON intrinsics"=17)) +
  labs(x=NULL, y="Speedup over NumPy", color=NULL, linetype=NULL, shape=NULL) +
  theme_paper +
  theme(legend.position="top", legend.key.size=unit(0.3,"cm"),
        legend.text=element_text(size=6), legend.margin=margin(0,0,-4,0),
        axis.text.x=element_text(angle=45, hjust=1, size=4),
        axis.text.y=element_text(size=5.5),
        axis.title.y=element_text(size=7),
        strip.text=element_text(face="bold", size=7),
        panel.border=element_rect(color="#cccccc", fill=NA, linewidth=0.3),
        panel.spacing=unit(0.5,"lines"))
```

\vspace{-0.8em}
\noindent\footnotesize\emph{Figure 1: Speedup of fused JIT kernels over NumPy. Gray line = 1x baseline (values below 1x clamped). Fused kernels sustain high speedup even at large N because fusion eliminates intermediate memory traffic.}
\normalsize
\vspace{0.3em}

Figure 1 shows the speedup for the three fused kernels. Unlike simple element-wise operations, these kernels maintain substantial speedups across all problem sizes.

\emph{Softmax} ($4.6\times$ at $N{=}4$M). NumPy computes softmax as four separate array passes: \texttt{max}, \texttt{subtract}, \texttt{exp}, and \texttt{divide(exp, sum)}. Each pass reads and writes the full array. Our JIT kernel fuses all passes into a single loop with stack-allocated accumulators (via \texttt{alloca} promoted to registers by LLVM), reducing memory traffic by ${\sim}4\times$.

\emph{Cross-entropy} ($5.6\times$ at $N{=}4$M). Similar to softmax but with an additional log-sum-exp computation. The fused kernel avoids materializing the softmax output entirely, computing the log-probability in a single pass.

\emph{RMSNorm} (${\sim}0.7\times$ at $N{=}4$M). RMSNorm computes $x / \sqrt{\mathrm{mean}(x^2) + \epsilon}$ in two passes: sum-of-squares then scale. NumPy's BLAS-accelerated vector operations (using Apple's Accelerate framework) outperform our single-threaded JIT at large $N$. The JIT kernel still achieves $1.9\times$ speedup up to $N{=}4$K where dispatch overhead dominates.

```{r fused-table, results='asis'}
ftbl <- df_1d %>%
  filter(kernel %in% c("softmax","cross_entropy","rmsnorm")) %>%
  group_by(kernel) %>%
  summarise(
    `Peak auto-vec`=sprintf("%.1fx (N=%s)", max(exo_speedup),
                            ifelse(n[which.max(exo_speedup)]>=1e3,
                                   paste0(n[which.max(exo_speedup)]/1e3,"K"),
                                   n[which.max(exo_speedup)])),
    `Peak NEON`=sprintf("%.1fx (N=%s)", max(neon_speedup),
                        ifelse(n[which.max(neon_speedup)]>=1e3,
                               paste0(n[which.max(neon_speedup)]/1e3,"K"),
                               n[which.max(neon_speedup)])),
    `Auto-vec @ 4M`=sprintf("%.1fx", exo_speedup[n==max(n)]),
    `NEON @ 4M`=sprintf("%.1fx", neon_speedup[n==max(n)]),
    .groups="drop"
  ) %>%
  mutate(kernel=c("softmax"="Softmax","cross_entropy"="Cross-Entropy","rmsnorm"="RMSNorm")[kernel])

cat("\\begin{table}[H]\n\\centering\n\\resizebox{\\columnwidth}{!}{")
knitr::kable(ftbl, format="latex", booktabs=TRUE, col.names=c("Kernel","Peak auto-vec","Peak NEON","Auto-vec @ 4M","NEON @ 4M"),
             align=c("l","r","r","r","r")) %>% cat()
cat("}\n\\caption{Fused kernel speedup over NumPy (Apple Silicon, float32). Peak speedup and sustained speedup at $N{=}4$M.}\n\\end{table}")
```

\textbf{Baseline Kernels.}

```{r baseline-plot, fig.width=3.3, fig.height=8.0, out.width="\\columnwidth"}
baselines <- sp %>%
  filter(kernel %in% c("saxpy","add","relu","dot","embedding")) %>%
  mutate(n_fct=factor(n, levels=sort(unique(n))),
         n_lab=ifelse(n>=1e6, paste0(n/1e6,"M"),
                ifelse(n>=1e3, paste0(n/1e3,"K"), as.character(n))))
n_labs2 <- baselines %>% distinct(n, n_lab) %>% arrange(n)

ggplot(baselines, aes(n_fct, speedup_clamped, color=variant, linetype=variant, group=variant)) +
  geom_hline(yintercept=1, color="#bbbbbb", linewidth=0.4) +
  geom_line(linewidth=0.5) +
  geom_point(aes(shape=variant), size=0.8) +
  facet_wrap(~kernel, scales="free_y", ncol=1) +
  scale_x_discrete(labels=setNames(n_labs2$n_lab, n_labs2$n)) +
  scale_color_manual(values=c("Auto-vectorized"="#4C72B0","NEON intrinsics"="#DD8452")) +
  scale_linetype_manual(values=c("Auto-vectorized"="solid","NEON intrinsics"="dashed")) +
  scale_shape_manual(values=c("Auto-vectorized"=16,"NEON intrinsics"=17)) +
  labs(x=NULL, y="Speedup over NumPy", color=NULL, linetype=NULL, shape=NULL) +
  theme_paper +
  theme(legend.position="top", legend.key.size=unit(0.3,"cm"),
        legend.text=element_text(size=5.5), legend.margin=margin(0,0,-4,0),
        axis.text.x=element_text(angle=45, hjust=1, size=3.5),
        axis.text.y=element_text(size=5),
        axis.title.y=element_text(size=6),
        strip.text=element_text(face="bold", size=6),
        panel.border=element_rect(color="#cccccc", fill=NA, linewidth=0.3),
        panel.spacing=unit(0.3,"lines"))
```

\vspace{-0.8em}
\noindent\footnotesize\emph{Figure 2: Baseline element-wise kernels. Speedup converges toward 1x at large N as both implementations become memory-bandwidth-bound.}
\normalsize
\vspace{0.3em}

Figure 2 shows results for the simple element-wise kernels. These confirm the expected pattern: at small $N$, reduced dispatch overhead yields $3{-}9\times$ speedups, but at large $N$ both implementations are memory-bandwidth-bound and speedups converge toward $1\times$. These kernels involve a single pass over the data with no opportunity for fusion.

For matrix kernels (matmul, matvec, weighted sum), NumPy's Accelerate-backed GEMM uses Apple's AMX coprocessor, reaching 2353~GFLOP/s at $n{=}512$ while our single-threaded JIT plateaus at ${\sim}40$~GFLOP/s. At small sizes ($n \leq 64$), the JIT achieves $3{-}5\times$ speedup due to BLAS dispatch overhead, but AMX hardware acceleration makes large-$N$ competition infeasible for single-core NEON.

\textbf{Auto-Vectorized vs.\ NEON Intrinsics.}

For the fused kernels, the auto-vectorized variant consistently outperforms hand-written NEON. LLVM's loop vectorizer, combined with fast-math flags and \texttt{noalias} pointer attributes, generates high-quality SIMD code from scalar loops and can apply global optimizations (SLP vectorization, instruction scheduling) across the entire fused loop body, something difficult to replicate with individual intrinsic operations.

The NEON variant is necessary only for patterns that LLVM cannot auto-vectorize, such as horizontal max reductions in softmax's max-finding pass. Here, explicit \texttt{neon\_fmax\_acc\_f32x4} with multiple accumulators for instruction-level parallelism is required. However, the overall fused kernel still performs better in auto-vectorized form because the non-reducible portion of the loop body benefits from LLVM's optimizer.

# Discussion

The central result is that kernel fusion, combining multiple array passes into a single loop, provides sustained speedups that do not vanish at large problem sizes. This is the key advantage a JIT compiler has over NumPy's array-at-a-time execution model.

\emph{Why fusion matters.} NumPy computes softmax as four separate operations: \texttt{max}, \texttt{subtract}, \texttt{exp}, and \texttt{divide}. Each reads and writes the full array, producing ${\sim}4\times$ the memory traffic of a fused implementation. At $N{=}4$M (${\sim}16$\,MB), the working set exceeds L1/L2 cache, making memory bandwidth the bottleneck. Fusion reduces this traffic to a single read-write pass, which is why softmax and cross-entropy sustain $4{-}6\times$ speedups even when simple kernels converge to $1\times$.

\emph{Stack allocation enables register promotion.} A custom \texttt{Stack} memory class uses \texttt{alloca} instead of \texttt{malloc} for scalar accumulators. LLVM promotes these to registers, avoiding memory round-trips in the inner loop. This is critical for fused kernels that maintain multiple running accumulators (max, sum, etc.).

\emph{Dispatch overhead is secondary.} The C extension with \texttt{vectorcall} achieves ${\sim}1\,\mu$s FFI overhead, which dominates at small $N$ but becomes negligible at large $N$. Fusion is the only mechanism that provides sustained large-$N$ speedups.

The primary limitation is that for simple element-wise operations and dense linear algebra, NumPy's Accelerate/AMX backend is faster. Supporting AMX or multi-threading would require extending the pipeline beyond its current scope of ${\sim}1{,}250$ lines of Python.

# Related Work

Halide [ragan2013halide] pioneered algorithm/schedule separation for image processing. Exo builds on this idea but externalizes hardware knowledge to user-defined libraries. TVM [chen2018tvm] extends the Halide model with learning-based auto-tuning. JAX [jax] provides trace-based JIT via XLA. Numba [numba] JIT-compiles Python via llvmlite through type inference but does not provide scheduling transforms. Our work is complementary: a JIT backend for a scheduling language (Exo), using a Python-native compiler stack (xDSL and llvmlite).

# Conclusion

We demonstrate that Exo kernels can be JIT-compiled to native ARM64 code through approximately 1,250 lines of Python, with no C compilation step. The key finding is that kernel fusion provides sustained speedups: softmax and cross-entropy maintain $4{-}6\times$ over NumPy even at $N{=}4$M, because fusing multiple passes into a single loop eliminates intermediate memory traffic. Simple element-wise kernels converge toward $1\times$ at large $N$ as expected, confirming that fusion is the essential mechanism for outperforming NumPy's array-at-a-time execution model. The pipeline runs entirely in-process and produces code competitive with LLVM's best auto-vectorization.

\footnotesize
\setlength{\parskip}{0.2ex}
\setlength{\parindent}{0pt}

# References {-}

[ikarashi2022exo] Y. Ikarashi, G. L. Bernstein, A. Reinking, H. Genc, and J. Ragan-Kelley. Exo: Exocompilation for productive programming of hardware accelerators. In *PLDI*, 2022.

[fehr2025xdsl] M. Fehr et al. xDSL: A common compiler ecosystem for languages that target dedicated accelerators. In *CGO*, 2025.

[numba2015llvmlite] Numba Development Team. llvmlite: A lightweight LLVM Python binding. \url{https://github.com/numba/llvmlite}.

[numba] S. K. Lam, A. Pitrou, and S. Seibert. Numba: A LLVM-based Python JIT compiler. In *LLVM-HPC Workshop*, 2015.

[ragan2013halide] J. Ragan-Kelley et al. Halide: A language and compiler for optimizing parallelism, locality, and recomputation in image processing pipelines. In *PLDI*, 2013.

[chen2018tvm] T. Chen et al. TVM: An automated end-to-end optimizing compiler for deep learning. In *OSDI*, 2018.

[jax] J. Bradbury et al. JAX: Composable transformations of Python+NumPy programs. \url{https://github.com/jax-ml/jax}.
